---
title: 'ESM 244 HW1A: Space Launch Visualization & Vegetable + Fruit PCA'
author: "Kynan Witters Hicks"
date: "January 22, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE}

library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(plyr)
library(factoextra)
library(ggbiplot)
library(MASS)
library(stringr)

```

#Part 1: Space Launch Visualization
##Load Data

```{r data, message=FALSE}

agencies <- read_csv("agencies.csv")
launches <- read_csv("launches.csv")

```

##Wrangle Launch Data

```{r select data to analyze}

state_names <- levels(launches$state_code)

launch_time <- launches %>% 
  group_by(launch_year, state_code) %>% 
  tally() %>% 
  dplyr::rename(launches = n) %>% 
  dplyr::rename(state = state_code) %>% 
  arrange(desc(launch_year), desc(launches)) %>% 
  mutate(name = case_when(
    grepl("J", state) ~ "Japan",
    grepl("US", state)~ "United States",
    grepl("BR", state)~ "Brazil",
    grepl("F", state)~ "France",
    grepl("CN", state)~ "China",
    grepl("I", state)~ "Israel",
    grepl("IL", state)~ "Italy",
    grepl("IR", state)~ "Iran",
    grepl("RU", state)~ "Russia",
    grepl("SU", state)~ "Soviet Union",
    grepl("UK", state)~ "United Kingdom",
    grepl("KP", state)~ "North Korea",
    grepl("KR", state)~ "South Korea",
    grepl("CYM", state)~ "Cayman Islands",
    grepl("I-ELDO", state)~ "Other",
    grepl("I-ESA", state)~ "Other")) 

```

##Visualize Space Launch Data 

```{r visualize, warning=FALSE}


#Extend color palette to include more colors. Doing it this way allows R to extrapolate colors out to 12 different colors rather than 9 (which is the origional number of colors in the "Spectral" palette).

colourCount = length(unique(launch_time$name))
getPalette = colorRampPalette(brewer.pal(12, "Spectral"))

#ggplot

gglaunch <- ggplot(launch_time, aes(x = launch_year, 
                                    y = launches,
                                    fill = name)) +
  geom_bar(stat="identity", width = 0.8) +
  scale_fill_manual(values = getPalette(colourCount)) +
  ylab("Number of Launches") +
  xlab("Year") +
  labs(title = "Number of Rocket Launches by Country from 1957 - 2018", 
       caption = "*Note: The Soviet Union was disbanded in 1991. \n Data from Economist GitHub - https://bit.ly/2TiCiiS.") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 12),
        axis.text = element_text(size = 10), 
        plot.caption = element_text(hjust = 0.5, 
                                    vjust = -1.0, 
                                    size = 10),
        legend.title = element_blank(),
        aspect.ratio = 0.5) +
  scale_x_continuous(breaks = c(seq(1955, 2025, by = 10)),
                     limits = c(1955, 2020)) +
  scale_y_continuous(breaks = c(seq(0, 160, by = 20)),
                     limits = c(0, 145))

gglaunch

```

#Part 2: Vegetable and Fruit PCA
##Load Data

```{r, message=FALSE}

###Part 2: Load Data

usda_nutrients <- read_csv("usda_nutrients.csv")

```

##Wrangle Data

```{r}

#### Part 2 - PCA Data Wrangle

veggy_fruit <- usda_nutrients %>% 
  filter(FoodGroup == "Vegetables and Vegetable Products"|
           FoodGroup == "Fruits and Fruit Juices") %>% 
  filter(str_detect(ShortDescrip,"RAW"))

```

##Conduct PCA

```{r}

### Part 2: PCA

nutrient_pca <- prcomp(veggy_fruit[9:30], scale = TRUE)

```
###PCA Results: 
1. 38.6% of variance is explained in the first two principle components. 
2. 83.9% of variance is explained by PC11 - good cutoff? 
3. The first two principal components explain the most variation in phosphorus and protein content while the second principal component is explained most by vitamin B and Selenium content 


##Visualize PCA

```{r}

fviz_pca_biplot(nutrient_pca,
                habillage = veggy_fruit$FoodGroup,
                palette = c("#FC4E07", "#E7B800"),
                addEllipses = TRUE,
                label = "var",
                repel = TRUE,
                title = "Biplot of PCA for Nutrients in Fruits and Vegetables") +
  theme_minimal() +
  theme(legend.position = 'bottom')
```

###Trends

1. Fruits and fruit juices don't have much protein, iron, or zinc in them, while many vegetables do. 
2. Surgar and vitamin C correlate to being present in many fruits and vegetables
3. Sugar and vitamin C are positively correlated while sugar is negatively correlated with protein, zinc, and iron. 
